---
title: "Week 6: Visualizing the Bayesian Workflow"
date: today
date-format: "DD/MM/YY"
format: pdf
execute: 
  warning: false
  message: false
---

# Introduction

This lab will be looking at trying to replicate some of the visualizations in the lecture notes, involving prior and posterior predictive checks, and LOO model comparisons.

The dataset is a 0.1% of all births in the US in 2017. I've pulled out a few different variables, but as in the lecture, we'll just focus on birth weight and gestational age.

# The data

Read it in, along with all our packages.

```{r}
library(tidyverse)
library(here)
# for bayes stuff
library(rstan)
library(bayesplot) 
library(loo) 
library(tidybayes) 

ds <- read_rds(here("data","births_2017_sample.RDS"))
head(ds)
```

Brief overview of variables:

-   `mager` mum's age
-   `mracehisp` mum's race/ethnicity see here for codes: https://data.nber.org/natality/2017/natl2017.pdf page 15
-   `meduc` mum's education see here for codes: https://data.nber.org/natality/2017/natl2017.pdf page 16
-   `bmi` mum's bmi
-   `sex` baby's sex
-   `combgest` gestational age in weeks
-   `dbwt` birth weight in kg
-   `ilive` alive at time of report y/n/ unsure

I'm going to rename some variables, remove any observations with missing gestational age or birth weight, restrict just to babies that were alive, and make a preterm variable.

```{r}
ds <- ds %>% 
  rename(birthweight = dbwt, gest = combgest) %>% 
  mutate(preterm = ifelse(gest<32, "Y", "N")) %>% 
  filter(ilive=="Y",gest< 99, birthweight<9.999)
```

## Question 1

Use plots or tables to show three interesting observations about the data. Remember:

-   Explain what your graph/ tables show
-   Choose a graph type that's appropriate to the data type
-   If you use `geom_smooth`, please also plot the underlying data

Feel free to replicate one of the scatter plots in the lectures as one of the interesting observations, as those form the basis of our models.

```{r}
ds |>
  ggplot(aes(x=log(gest), y=log(birthweight), color=preterm)) +
  geom_point()+
  geom_smooth(method = "lm")
  
```

I plotted the scatter plot of log(birthweight) vs log(gestational age) which forms the basis of the model. I also use different color scales to indicate whether the birth was preterm or not. I found that log(birthweight) increases as log(gestational age) increases, and the increase is faster (slope is steeper) in preterm birth.

```{r}
ds |>
  ggplot(aes(x=bmi, y=birthweight)) +
  geom_point()
  
```
I plotted a scatter plot of birthweight vs bmi. Overall, there is no clear pattern between these two variables. On a side note, a bmi value close to 100 seems to be an outlier.
```{r}
ds |>
  ggplot(aes(x=as.factor(mracehisp), y=birthweight))+
  geom_boxplot()
```
I plotted boxplots of birthweight by mom's race, and found that race 5 (Non-Hispanic NHOPI (only) ) has denser concentration around birthweight 3kg while other races' birthweight are more spread out. Race 1 (Non-Hispanic White (only)), 2 (Non-Hispanic Black (only) ) and 7 (Hispanic) have more outlier birthweights than other races. 

# The model

As in lecture, we will look at two candidate models

Model 1 has log birth weight as a function of log gestational age

$$
\log(y_i) \sim N(\beta_1 + \beta_2\log(x_i), \sigma^2)
$$

Model 2 has an interaction term between gestation and prematurity

$$
\log(y_i) \sim N(\beta_1 + \beta_2\log(x_i) + \beta_2 z_i + \beta_3\log(x_i) z_i, \sigma^2)
$$

-   $y_i$ is weight in kg
-   $x_i$ is gestational age in weeks, CENTERED AND STANDARDIZED
-   $z_i$ is preterm (0 or 1, if gestational age is less than 32 weeks)

# Prior predictive checks

Let's put some weakly informative priors on all parameters i.e. for the $\beta$s

$$
\beta \sim N(0, 1)
$$

and for $\sigma$

$$
\sigma \sim N^+(0,1)
$$ where the plus means positive values only i.e. Half Normal.

Let's check to see what the resulting distribution of birth weights look like given Model 1 and the priors specified above, assuming we had no data on birth weight (but observations of gestational age).

## Question 2

For Model 1, simulate values of $\beta$s and $\sigma$ based on the priors above. Do 1000 simulations. Use these values to simulate (log) birth weights from the likelihood specified in Model 1, based on the set of observed gestational weights. **Remember the gestational weights should be centered and standardized**.

-   Plot the resulting distribution of simulated (log) birth weights.
-   Plot ten simulations of (log) birthweights against gestational age.


```{r}
# code inspired from https://www.monicaalexander.com/posts/2020-28-02-bayes_viz/
set.seed(1)
nsims=1000
sigma <- abs(rnorm(nsims, 0, 1))
beta1 <- rnorm(nsims, 0, 1)
beta2 <- rnorm(nsims, 0, 1)

dsims <- tibble(log_gest_c = (log(ds$gest)-mean(log(ds$gest)))/sd(log(ds$gest)))

for(i in 1:nsims){
  this_mu <- beta1[i] + beta2[i]*dsims$log_gest_c 
  dsims[paste0(i)] <- this_mu + rnorm(nrow(dsims), 0, sigma[i])
}

dsl <- dsims %>% 
  pivot_longer(`1`:`1000`, names_to = "sim", values_to = "sim_weight")

dsl %>% 
  ggplot(aes(sim_weight)) + geom_histogram(aes(y = ..density..), bins = 20, fill = "turquoise", color = "black") + 
  theme_bw(base_size = 16) 
```
```{r}
dsl_10 <- dsims[,1:11] |>
  pivot_longer(`1`:`10`, names_to = "sim", values_to = "sim_weight")

dsl_10|>
  ggplot(aes(x=log_gest_c, y=sim_weight,color=sim))+
  geom_point()
```

# Run the model

Now we're going to run Model 1 in Stan. The stan code is in the `code/models` folder.

First, get our data into right form for input into stan.

```{r}
ds$log_weight <- log(ds$birthweight)
ds$log_gest_c <- (log(ds$gest) - mean(log(ds$gest)))/sd(log(ds$gest))

# put into a list
stan_data <- list(N = nrow(ds),
                  log_weight = ds$log_weight,
                  log_gest = ds$log_gest_c)
```

Now fit the model

```{r}
mod1 <- stan(data = stan_data, 
             file = here("code/models/simple_weight.stan"),
             iter = 500,
             seed = 243)
```

```{r}
summary(mod1)$summary[c("beta[1]", "beta[2]", "sigma"),]
```

## Question 3

Based on model 3, give an estimate of the expected birthweight of a baby who was born at a gestational age of 37 weeks.
```{r}
exp(1.1626250+0.1436183*(log(37) - mean(log(ds$gest)))/sd(log(ds$gest)))
```
The expected birthweight is 2.93654 kg.

## Question 4

Write a stan model to run Model 2, and run it.
```{r}
ds$log_weight <- log(ds$birthweight)
ds$log_gest_c <- (log(ds$gest) - mean(log(ds$gest)))/sd(log(ds$gest))
ds$prematurity <- ifelse(ds$preterm=="Y", 1, 0)
# put into a list
stan_data <- list(N = nrow(ds),
                  log_weight = ds$log_weight,
                  log_gest = ds$log_gest_c,
                  prematurity = ds$prematurity,
                  combo = ds$prematurity*ds$log_gest_c)
```
```{r}
# model is beta1+beta3*log_gest+beta2*premature+beta4*log_gest*premature
mod2 <- stan(data = stan_data, 
             file = here("code/models/simple_weight_mod2.stan"),
             iter = 500,
             seed = 243)
```
```{r}
summary(mod2)$summary[c("beta[1]", "beta[2]", "beta[3]", "beta[4]", "sigma"),]
```

## Question 5

For reference I have uploaded some model 2 results. Check your results are similar.

```{r}
load(here("output", "mod2.Rda"))
summary(mod2)$summary[c(paste0("beta[", 1:4, "]"), "sigma"),]
```

# PPCs

Now we've run two candidate models let's do some posterior predictive checks. The `bayesplot` package has a lot of inbuilt graphing functions to do this. For example, let's plot the distribution of our data (y) against 100 different datasets drawn from the posterior predictive distribution:

```{r}
set.seed(1856)
y <- ds$log_weight
yrep1 <- extract(mod1)[["log_weight_rep"]]
yrep2 <- extract(mod2)[["log_weight_rep"]] 
dim(yrep1)
samp100 <- sample(nrow(yrep1), 100)
ppc_dens_overlay(y, yrep1[samp100, ])  + ggtitle("distribution of observed versus predicted birthweights")
```

## Question 6

Make a similar plot to the one above but for model 2, and **not** using the bayes plot in built function (i.e. do it yourself just with `geom_density`)
```{r}
# code inspired from https://www.monicaalexander.com/posts/2020-28-02-bayes_viz/
# first, get into a tibble
rownames(yrep1) <- 1:nrow(yrep1)
dr <- as_tibble(t(yrep1))
dr <- dr %>% bind_cols(i = 1:length(ds$birthweight), log_weight_obs = log(ds$birthweight))

# turn into long format; easier to plot
dr <- dr %>% 
  pivot_longer(-(i:log_weight_obs), names_to = "sim", values_to ="y_rep")

# filter to just include 100 draws and plot!
dr %>% 
  filter(sim %in% samp100) %>% 
  ggplot(aes(y_rep, group = sim)) + 
  geom_density(alpha = 0.2, aes(color = "y_rep")) + 
  geom_density(data = ds %>% mutate(sim = 1), 
               aes(x = log(birthweight), col = "y")) + 
  scale_color_manual(name = "", 
                     values = c("y" = "darkblue", 
                                "y_rep" = "lightblue")) + 
  ggtitle("Distribution of observed and replicated birthweights") + 
  theme_bw(base_size = 12)
```

## Test statistics

We can also look at some summary statistics in the PPD versus the data, again either using `bayesplot` -- the function of interest is `ppc_stat` or `ppc_stat_grouped` -- or just doing it ourselves using ggplot.

E.g. medians by prematurity for Model 1

```{r}
ppc_stat_grouped(ds$log_weight, yrep1, group = ds$preterm, stat = 'median')
```

## Question 7

Use a test statistic of the proportion of births under 2.5kg. Calculate the test statistic for the data, and the posterior predictive samples for both models, and plot the comparison (one plot per model).
```{r}
# code inspired from https://www.monicaalexander.com/posts/2020-28-02-bayes_viz/
t_y <- mean(y<=log(2.5))
t_y_rep <- sapply(1:nrow(yrep1), function(i) mean(yrep1[i,]<=log(2.5)))
t_y_rep_2 <- sapply(1:nrow(yrep2), function(i) mean(yrep2[i,]<=log(2.5)))

ggplot(data = as_tibble(t_y_rep), aes(value)) + 
    geom_histogram(aes(fill = "replicated")) + 
    geom_vline(aes(xintercept = t_y, color = "observed"), lwd = 1.5) + 
  ggtitle("Model 1: proportion of births less than 2.5kg") + 
  theme_bw(base_size = 12) + 
  scale_color_manual(name = "", 
                     values = c("observed" = "darkblue"))+
  scale_fill_manual(name = "", 
                     values = c("replicated" = "lightblue")) 

ggplot(data = as_tibble(t_y_rep_2), aes(value)) + 
    geom_histogram(aes(fill = "replicated")) + 
    geom_vline(aes(xintercept = t_y, color = "observed"), lwd = 1.5) + 
  ggtitle("Model 2: proportion of births less than 2.5kg") + 
  theme_bw(base_size = 12) + 
  scale_color_manual(name = "", 
                     values = c("observed" = "darkblue"))+
  scale_fill_manual(name = "", 
                     values = c("replicated" = "lightblue")) 
```
```{r}
t_y
```
The average of the proportion of births under 2.5kg is 0.08198855. 

# LOO

Finally let's calculate the LOO elpd for each model and compare. The first step of this is to get the point-wise log likelihood estimates from each model:

```{r}
loglik1 <- extract(mod1)[["log_lik"]]
loglik2 <- extract(mod2)[["log_lik"]]
```

And then we can use these in the `loo` function to get estimates for the elpd. Note the `save_psis = TRUE` argument saves the calculation for each simulated draw, which is needed for the LOO-PIT calculation below.

```{r}
loo1 <- loo(loglik1, save_psis = TRUE)
loo2 <- loo(loglik2, save_psis = TRUE)
```

Look at the output:

```{r}
loo1
loo2
```

Comparing the two models tells us Model 2 is better:

```{r}
loo_compare(loo1, loo2)
```

We can also compare the LOO-PIT of each of the models to standard uniforms. The both do pretty well.

```{r}
ppc_loo_pit_overlay(yrep = yrep1, y = y, lw = weights(loo1$psis_object))
ppc_loo_pit_overlay(yrep = yrep2, y = y, lw = weights(loo2$psis_object))
```

## Bonus question (not required)

Create your own PIT histogram "from scratch" for Model 2.

## Question 8

Based on the original dataset, choose one (or more) additional covariates to add to the linear regression model. Run the model in Stan, and compare with Model 2 above on at least 2 posterior predictive checks.

I add (centered and standardized) (log) bmi to model2.  
The model is beta1+beta3*log_gest+beta2*premature+beta4*log_gest*premature+beta5*log_bmi.
```{r}
ds$log_weight <- log(ds$birthweight)
ds$log_gest_c <- (log(ds$gest) - mean(log(ds$gest)))/sd(log(ds$gest))
ds$prematurity <- ifelse(ds$preterm=="Y", 1, 0)
ds$log_bmi_c <- (log(ds$bmi) - mean(log(ds$bmi)))/sd(log(ds$bmi))
# put into a list
stan_data <- list(N = nrow(ds),
                  log_weight = ds$log_weight,
                  log_gest = ds$log_gest_c,
                  prematurity = ds$prematurity,
                  combo = ds$prematurity*ds$log_gest_c,
                  log_bmi = ds$log_bmi_c)
```
```{r}

mod3 <- stan(data = stan_data, 
             file = here("code/models/simple_weight_mod_q8.stan"),
             iter = 500,
             seed = 243)
```
```{r}
summary(mod3)$summary[c("beta[1]", "beta[2]", "beta[3]", "beta[4]", "beta[5]", "sigma"),]
```
Then I checked the median by prematurity and proportion of births under 2.5kg.
```{r}
set.seed(1856)
y <- ds$log_weight
yrep3 <- extract(mod3)[["log_weight_rep"]]
```
```{r}
#model3
ppc_stat_grouped(ds$log_weight, yrep3, group = ds$preterm, stat = 'median')

#model2
ppc_stat_grouped(ds$log_weight, yrep2, group = ds$preterm, stat = 'median')
```
```{r}
# code taken from https://www.monicaalexander.com/posts/2020-28-02-bayes_viz/
t_y <- mean(y<=log(2.5))
t_y_rep_3 <- sapply(1:nrow(yrep3), function(i) mean(yrep3[i,]<=log(2.5)))
t_y_rep_2 <- sapply(1:nrow(yrep2), function(i) mean(yrep2[i,]<=log(2.5)))

ggplot(data = as_tibble(t_y_rep_3), aes(value)) + 
    geom_histogram(aes(fill = "replicated")) + 
    geom_vline(aes(xintercept = t_y, color = "observed"), lwd = 1.5) + 
  ggtitle("Model 3: proportion of births less than 2.5kg") + 
  theme_bw(base_size = 12) + 
  scale_color_manual(name = "", 
                     values = c("observed" = "darkblue"))+
  scale_fill_manual(name = "", 
                     values = c("replicated" = "lightblue")) 

ggplot(data = as_tibble(t_y_rep_2), aes(value)) + 
    geom_histogram(aes(fill = "replicated")) + 
    geom_vline(aes(xintercept = t_y, color = "observed"), lwd = 1.5) + 
  ggtitle("Model 2: proportion of births less than 2.5kg") + 
  theme_bw(base_size = 12) + 
  scale_color_manual(name = "", 
                     values = c("observed" = "darkblue"))+
  scale_fill_manual(name = "", 
                     values = c("replicated" = "lightblue")) 
```
For both median by preterm and proportion of births under 2.5kg, the replicated values are both around the observed. There is no significant difference between model3 (add variable) and model2.